On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-30 tex/PaperI/draft4.tex
more abstract pruning

% !TEX TS-program = pdflatexmk

%\documentstyle[aas2pp4,epsf]{article}
%%%\documentstyle[aaspp4,epsf]{article}
%\documentstyle[12pt,aasms]{article}    % this is for a preprint
%(single-spaced)
%\documentstyle[aaspp4,epsf]{article} % this is for small print
%\documentstyle[12pt, aaspp4]{article}

%\documentstyle[11pt,aaspp]{article}
%\documentclass[12pt, preprint]{aastex}

%\documentclass[manuscript]{aastex}
\documentclass[apj, numberedappendix]{emulateapj}

%\documentclass[12pt, preprint,numberedappendix]{emulateapj}
%\documentstyle[12pt,aasms]{article}    % this is for submittal
                                       % (double-spaced)

%\documentstyle[12pt,aasms]{article}   \usepackage{emulateapj5}

\usepackage{graphicx} 
%\usepackage{graphics}                       
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{subeqnarray}
\usepackage{epstopdf}
%\bibliographystyle{apj}

@@

\newcommand{\gcc}{\;\mathrm{g\; cm^{-3}}}
\newcommand{\gsc}{\;\mathrm{g\; cm^{-2}}}
\newcommand{\cm}{\; {\rm cm}}
\newcommand{\mm}{\; {\rm mm}}
%\newcommand{\ps}{\; {\rm s^{-1}}}
\newcommand{\km}{\; {\rm km}}
%\newcommand{\au}{\; \varpi_{\rm AU}}

@@

\newcommand{\delad}{\nabla_{\rm ad}}
\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\RH}{R_{\rm H}}
\newcommand{\co}{_{\rm c}}
\newcommand{\pla}{_{\rm p}}
\newcommand{\di}{_{\rm d}}
\newcommand{\cb}{_{\rm RCB}}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1} }
\newcommand{\mpn}[1] { m_{ \rm p #1} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1} }
%\newcommand{\aun}[1]{ a_{#1\oplus} }
\def\crit{_{\rm{c, crit}}}

@@

The core accretion hypothesis proposes that giant planets form by gas accretion onto a solid protoplanetary core. The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. This value, however, depends on several factors, including the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.  Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk. Our model investigates atmospheric growth around solid cores of fixed mass.  Without planetesimal accretion heating the envelope, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction. We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the disk.  For different disk locations, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass is $\sim$$8.5 M_{\oplus}$ at 5 AU and decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard dust opacity.  This trend is mainly explained by the cooler temperatures in the outer disk, while disk pressures have a negligible effect. At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the gas reduces the critical core mass. We also develop a non-self-gravitating, analytic cooling model, which reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$10\%$ as massive as the core.

>>>>

\documentclass[apj, numberedappendix]{emulateapj}

\usepackage{graphicx} 
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{subeqnarray}
\usepackage{epstopdf}
%\bibliographystyle{apj}

@@

\newcommand{\gcc}{\;\mathrm{g\; cm^{-3}}}
\newcommand{\gsc}{\;\mathrm{g\; cm^{-2}}}
\newcommand{\cm}{\; {\rm cm}}
\newcommand{\mm}{\; {\rm mm}}
%\newcommand{\ps}{\; {\rm s^{-1}}}
\newcommand{\km}{\; {\rm km}}

@@

\newcommand{\delad}{\nabla_{\rm ad}}
\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\RH}{R_{\rm H}}
\newcommand{\co}{_{\rm c}}
\newcommand{\pla}{_{\rm p}}
\newcommand{\di}{_{\rm d}}
\newcommand{\cb}{_{\rm RCB}}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1} }
\newcommand{\mpn}[1] { m_{ \rm p #1} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1} }
\def\crit{_{\rm{c, crit}}}

@@

In the core accretion hypothesis, giant planets form by gas accretion onto solid protoplanetary cores.  While typically quoted as  $10 M_\oplus$, the minimum (or critical) core mass to form a gas giant varies, depending on several factors: location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.  Motivated by ongoing direct imaging searches for giant planets, this study investigates core mass requirements in the outer disk.  To determine the fastest allowed rates of gas accretion, we consider solid cores that no longer accrete planetesimals, as this would heat the gas envelope. 
Our spherical, two-layer atmospheric cooling model includes an inner convective region and an outer radiative zone, which matches onto the disk.  We determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard dust opacities.  Cooler temperatures in the outer disk explain this trend, while variations in disk density are less influential.  At all distances, a lower dust opacity or higher mean molecular weight reduces the critical core mass. Our non-self-gravitating, analytic cooling model reveals that self-gravity significantly affects early atmospheric evolution, starting when the atmosphere is only $\sim$$10\%$ as massive as the core.
